---
title: "RepData_PeerAssessment1"
author: "Viswathish Dinesh"
date: "`r Sys.Date()`"
output: html_document
---

## Loading required packages
```{r, results=FALSE}
library(stats)
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

### Loading the data into R:
```{r Loading}
unzip("activity.zip")
rawdata <- read.csv("activity.csv")
```

### Convert date column into date format, view summary of data:
```{r PreProcessing}
rawdata$date <- as.Date(rawdata$date, format = "%Y-%m-%d")
```

### Grouping data by different parameters
```{r Grouping}
by_date <- group_by(rawdata, date)
date_stats <- summarize(by_date, total = sum(steps, na.rm = T),
                    mean = mean(steps, na.rm = T),
                    median = median(steps, na.rm = T))

by_int <- group_by(rawdata, interval)
int_stats <- summarize(by_int, total = sum(steps, na.rm = T),
                    mean = mean(steps, na.rm = T),
                    median = median(steps, na.rm = T))
```

## What is the mean total number of steps taken per day?

### Histogram of total steps per day
```{r totalHist}
ggplot(date_stats, aes(date, total)) + 
        geom_bar(stat = "identity", aes(fill = total)) + 
        labs(title = "Total Steps per Day") + 
        xlab(label = "Date") + 
        ylab(label = "Total Steps")
```

### Histogram of mean and median steps per day
```{r mean&median}
ggplot(date_stats, aes(date, mean)) + 
        geom_bar(stat = "identity", aes(fill = total)) + 
        labs(title = "Average Steps per Day (Mean)") + 
        xlab(label = "Date") + 
        ylab(label = "Average Steps")
ggplot(date_stats, aes(date, median)) + 
        geom_bar(stat = "identity", aes(fill = total)) + 
        labs(title = "Average Steps per Day (Median)") + 
        xlab(label = "Date") + 
        ylab(label = "Average Steps")
```

### Calculating mean and median of total steps per day
```{r mean$median total}
tot_mean <- round(mean(date_stats$total), digits = 2)
tot_median <- median(date_stats$total)
```
The mean number of total steps per day is `r tot_mean`, and the median is `r tot_median`.

## What is the average daily activity pattern?

### Which 5-minute interval, on average across all days, has the maximum amount of steps?
```{r max daily activity}
max_steps <- which.max(int_stats$mean)
max_interval <- int_stats[max_steps,]
```
The 5-minute interval with the maximum amount of steps, on average, is `r max_interval$interval`.

### Time series of average number of steps throughout the day
```{r daily activity}
ggplot(int_stats, aes(interval,mean)) + 
        geom_line(linewidth = 1) +
        xlab(label = "Time Interval") +
        ylab(label = "Average Number of Steps") +
        labs(title = "Average Number of Steps per 5-minute Time Interval")
```


## Imputing Missing Values

### Calculating the total number of missing values
```{r how many missing}
na_values <- sum(as.integer(is.na(rawdata$steps)))
```
There are 'r na_values' missing values.  
Let's use the average of each time interval to replace the missing values.

### Creating a new data set that replaces NA values of original data set
```{r imputing data}
filled_data <- rawdata
filled_data$steps <- ifelse(is.na(filled_data$steps),
                            yes = int_stats$mean[int_stats$interval %in% filled_data$interval],
                            no = filled_data$steps)
```

### Generating a histogram of total number of steps per day
```{r imputed histogram}
imp_by_date <- group_by(filled_data, date)
imputed_stats <- summarize(imp_by_date, total = sum(steps),
                    mean = mean(steps),
                    median = median(steps))
ggplot(imputed_stats, aes(date, total)) + 
        geom_bar(stat = "identity", aes(fill = total)) + 
        labs(title = "Total Steps per Day") + 
        xlab(label = "Date") + 
        ylab(label = "Total Steps")
```

### Calculating mean and median total steps per day, with imputed data
```{r mean&median total, imputed}
tot_mean_imp <- round(mean(imputed_stats$total), digits = 2)
tot_median_imp <- round(median(imputed_stats$total), digits = 2)
```
The mean number of total steps per day is `r as.character(tot_mean_imp)`, and the median is `r as.character(tot_median_imp)`. These values differ from the initial estimates, and the effect of imputing the data for the NA values has made the mean and median the same.

## Differences in activity patterns between weekdays and weekends

### Adding factor column to filled data
```{r adding day type}
filled_data$day_type <- ifelse(weekdays(filled_data$date) %in% c("Saturday", "Sunday"),
                               yes = "Weekend",
                               no = "Weekday")
filled_data$day_type <- as.factor(filled_data$day_type)
```

### Creating panel plot of average steps per 5-minute interval
```{r Wday vs Wend plots}
by_int_imp <- group_by(filled_data, interval, day_type)
int_stats_imp <- summarize(by_int_imp, total = sum(steps),
                    mean = mean(steps),
                    median = median(steps),
                    day_type = day_type)
ggplot(int_stats_imp, aes(interval,mean)) + 
        facet_grid(day_type~.) +
        geom_line(linewidth = 1) +
        xlab(label = "Time Interval") +
        ylab(label = "Average Number of Steps") +
        labs(title = "Average Number of Steps per 5-minute Time Interval")
```

During weekdays, the majority of steps taken are in the earlier part of the day, while the distribution of steps during weekends are more spread out during the day. Steps also begin earlier in the day overall on weekdays compared to weekends.